Neural models for computer-assisted musical orchestration: a preliminary study

### 2020-05-13 paper/latex/paper.tex
Revise CNN section

Our idea is to perform orchestration using deep neural networks trained for classification. In our setting, each class is an instrument-note pair. Our model takes in the mel spectrogram of a soundfile that is a combination of different instruments and pitches, and attempts to identify which instrument-pitch pairs are present. The network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. Then, when a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. By taking the samples that have the highest probability, an orchestrated solution can be created.

@@

The first deep model we tested was a Convolutional Neural Network (CNN). CNNs show good performance on classification problems for their ability to extract spatial features. \textbf{Reference needed}. Our architecture contained three convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a $2 \times 2$  MaxPool layer with a stride of 2. The kernel size is $3 \times 3$ with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32.

@@

Given the sequential nature of audio, a Long Short-Term Memory (LSTM) network provides a way to learn long term dependencies in the data \cite{LSTM}.

 In our setting, we added one LSTM layer with 32 outputs units following three convolutional layers following the same pattern as the CNN described earlier. After LSTM layer, we reshaped the output into the same size as the output of the last convulational layer, which was fed to another convolutional layer with the same structure and kernel size, yielding a tensor of dimensions $32\times 8 \times 21$. Finally, we flattened out the outputs and fed them into fully a connected layer with Dropout layer, followed by another fully connected layer. Finally, the sigmoid function was applied to the final layer in order to get a probability between 0 and 1 for each individual label. Since there is no competition relationship between different classes for each class can be classification result equally, it is reasonable for us to take sigmoid activations and used binary classification for each class to treat them independently. We minimize the binary cross entropy(BCE) loss between the predictions and targets. We implemented our neural networks with PyTorch and trained them with Adam optimizer. The model architecture is shown in Figure\ref{cnn_lstm}.\\

@@

To make our setting more reasonable, we took the well-known deep residual network(ResNet) as backbone in our experiments. Specifically, we used 18-layer ResNet, which allows information to pass directly from the input to the final layer of each block. Besides, to make the model more suitable to our problem, we reduce the number of parameters and reset the output channel numbers of each block with 32, 64, 32, 32 respectively. We took the same loss function-BCE and optimizer-Adam as those in training the CNN with LSTM model.

@@

During training, the loss function used to optimize the inner parameters of the model was binary cross entropy, as it is the common choice for multiclass multilabel classification frameworks such as this one. However, the value of the loss function alone is difficult to interpret. This is why, we used a complementary function $f$ only for evaluation, to match the notion of accuracy of the model with the intuitive approach. This function compares a vector of expected outputs $\overline{X}$ with the estimated output from the model $\hat{X}$ by using the following function

@@

\begin{equation}
M_N(\hat{X})_i = \left\{\begin{array}{ll}
1 \text{ if } i \in I_N(\hat{X})\\
0 \text{ otherwise}
\end{array}\right.
\label{NMax}
\end{equation}
and $I_N(\hat{X})$ is the set of the indices containing the $N$ first maximums of the vector $\hat{X}$.

More specifically, the function $M_N(\hat{X})$ takes as an input a vector of probabilities (values between 0 and 1) and outputs a vector where only the positions of the $N$ first maxima are set to 1. This new vector would be the orchestration of $N$ instruments given by the model. Thus, the function $f$ simply outputs the proportion of the estimated orchestration that matches the expected one.

@@

\section{Orchestration experiments}
- set of 10 targets
- orchestra fix to 10
- qualitative evaluation: acoustic inspection of the solutions
- quantitative evaluation: distance in feature space
- comparison table between Orchidea and our model

@@

\begin{table}
  \begin{center}
    \label{orch_eval}
    \begin{tabular}{|c|c|} 
    	  \hline
      \textbf{Model} & \textbf{Average Distance} \\
      \hline
      CNN with LSTM &   \\
      \hline
      ResNet &  \\
      \hline
      OrchIdea &  \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Comparison of the orchestration solutions of our two models, CNN with LSTM and ResNet, against OrchIdea's solutions. The distance between the target and solution are calculated for each target using \eqref{distance} and then averaged.}
\end{table}

@@

\begin{figure}
\includegraphics[scale=0.6]{figs/CNN4.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN5.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN6.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN7.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN8.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN9.png}
\end{figure}
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN10.png}
\end{figure}

>>>>

Our idea is to perform orchestration using deep neural networks trained for classification. In our setting, each class is an instrument-note pair. Our model takes in the mel spectrogram of a soundfile that is a combination of different instruments and pitches, and attempts to identify which instrument-pitch pairs are present. The network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. Then, when a network is given a target sound as input, it will attempt to apply the same classification rules, outputting a vector of the probabilities of each sample being in the target sound. By taking the samples that have the highest probability, an orchestrated solution can be created. Both of our deep models use the Adam optimizer, Binary Cross-Entropy loss, and were written using PyTorch.

@@

The first deep model we tested was a Convolutional Neural Network (CNN) with a Long Short-Term Memory (LSTM) unit. CNNs show good performance on classification problems for their ability to extract spatial features. \textbf{Reference needed}. Long Short-Term Memory (LSTM) units provide a way to learn long term dependencies in the data \cite{LSTM}, which is relevent given the sequential nature of audio.

Our architecture contains four convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a $2 \times 2$  MaxPool layer with a stride of 2. The kernel size is $3 \times 3$ with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32, and 32.

Following the first three convolutional layers, there is an LSTM layer with 32 outputs. After the LSTM layer, there is a final convolutional layer yielding a tensor of dimensions $32\times 8 \times 21$. We flatten the outputs and feed them into fully a connected layer with Dropout, then another another fully connected layer. Finally, the sigmoid function is applied to the final layer.

Since each class is independent, we are able to take the sigmoid activation and use binary classification for each class. The model architecture is shown in Figure \ref{cnn_lstm}.

@@

@@

To make our setting more reasonable, we took the well-known deep residual network (ResNet) as backbone in our experiments. Specifically, we used 18-layer ResNet, which allows information to pass directly from the input to the final layer of each block. Besides, to make the model more suitable to our problem, we reduce the number of parameters and reset the output channel numbers of each block with 32, 64, 32, 32 respectively.

@@

During training, the loss function used to optimize the inner parameters of the model was binary cross entropy, as it is the common choice for multiclass multilabel classification frameworks such as this one. However, the value of the loss function alone is difficult to interpret. For this reason we created a complementary function $f$ to be used only for evaluation. This function compares a vector of expected outputs $\overline{X}$ with the estimated output from the model $\hat{X}$ by using the following function

@@

\begin{equation}
M_N(\hat{X})_i = \left\{\begin{array}{ll}
1 \text{ if } i \in I_N(\hat{X})\\
0 \text{ otherwise}
\end{array}\right.
\label{NMax}
\end{equation}
and $I_N(\hat{X})$ is the set of indices containing the $N$ first maximums of the vector $\hat{X}$.

More specifically, the function $M_N(\hat{X})$ takes as an input a vector of probabilities and outputs a vector where only the positions of the $N$ first maxima are set to 1. This new vector would be the orchestration of $N$ instruments given by the model. Thus, the function $f$ simply outputs the proportion of the estimated orchestration that matches the expected one.

@@

\section{Orchestration experiments}

@@

\begin{table}
  \begin{center}
    \label{orch_eval}
    \begin{tabular}{|c|c|} 
    	  \hline
      \textbf{Model} & \textbf{Average Distance} \\
      \hline
      CNN with LSTM &   \\
      \hline
      ResNet &  \\
      \hline
      OrchIdea &  \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Comparison of the orchestration solutions of our two models, CNN with LSTM and ResNet, against OrchIdea's solutions. The distance between the target and solution are calculated for each target using Eqn. \eqref{distance} and then averaged.}
\end{table}

@@

\begin{figure}
\includegraphics[scale=0.6]{figs/CNN10.png}
\end{figure}
